---
title: "Sparse Canonical Correlation Analysis (SCCA) with the package flashpcaR"
author: "Gad Abraham, Rodrigo Canovas"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
csl: biomed-central.csl
bibliography: bibliography.bib
vignette: >
   %\VignetteIndexEntry{flashpcaR}
   %\VignetteEngine{knitr::rmarkdown}
   %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

The package flashpcaR implements sparse canonical correlation analysis (SCCA)
[@Parkhomenko2009;@Witten2009;@Witten2009b;@Witten2009c].

Assume we have two datasets $\mathbf{X}$ ($n \times m$ matrix) and $\mathbf{Y}$ ($n \times k$ matrix),
where the two matrices have been standardised such that each column has zero-mean and unit-variance.
Sparse canonical correlation analysis is essentially a low-rank ($D$-dimension) decomposition of the covariance of
$$\mathbf{\Sigma}_{XY}=\frac{1}{n-1}\mathbf{X}^T \mathbf{Y} \approx \mathbf{U}\mathbf{D}\mathbf{V}^T,$$
subject to constraints $\Vert \mathbf{U} \Vert_2 = 1, \Vert \mathbf{U}_j \Vert_1<s_1$ for $j=1,\ldots,d$
and $\Vert \mathbf{V}_2 \Vert = 1, \Vert \mathbf{V}_j \Vert_1<s_2$ for $j=1,\ldots,d$,
where $\Vert \cdot \Vert_2$ and $\Vert \cdot \Vert_1$ are
the $\ell_2$-norm and $\ell_1$-norm of the vector, respectively
(i.e., root of sum of square values and sum of absolute values, resp.).

In other words, we seek to find the canonical vectors $\mathbf{U}$ and $\mathbf{V}$ that maximise the correlation
$$\mbox{cor}(\mathbf{X} \mathbf{U}_d, \mathbf{Y} \mathbf{V}_d)$$
for dimensions $d=1,...,D$ (subject to the above constraints).

In practice, instead of the constraint formulation, we use a penalised formulation, where we penalise
the canonical vectors by penalties $\lambda_1 \geq 0$ and $\lambda_2 \geq 0$ [@Parkhomenko2009],
which is achieved by soft-thresholding
$$S(u_j, \lambda_1) = \mbox{sgn}(u_j) \max(0, \left| u_j \right| - \lambda_1)$$
and then subsequent renormalisation $\mathbf{u}^*=\frac{\mathbf{u}}{\Vert \mathbf{u} \Vert_2}$.
The penalties are tuned via cross-validation, such that we maximise the cross-validated
correlation above.

## Outline of the analysis

The standard SCCA pipeline is:

1. Prepare the two datasets, including any quality control needed
2. Perform SCCA in cross-validation over a sensible grid of penalties
3. Inspect the results; if needed repeat step 2 with different penalties.
4. One satisfied with cross-validation results, use the optimal penalties to train
one model over the entire data.
5. Inspect the results for the final model.

## Example

This example is modified from the OmicsPLS package (https://github.com/selbouhaddani/OmicsPLS)
[@el2018integrating], using the DILGOM gene expression and metabolite level data
(n=512 individuals) [@inouye2010immune;@inouye2010metabonomic]. The original gene expression data are at http://www.ebi.ac.uk/arrayexpress/files/E-TABM-1036
and the metabolite data are at https://www.embopress.org/action/downloadSupplement?doi=10.1038%2Fmsb.2010.93&file=msb201093-sup-0002.zip.
Pre-processing has been performed in the script `rna_metab_data.R` available
at https://github.com/gabraham/flashpca/blob/master/data/DILGOM.

The preprocessed datasets can be downloaded via:
```{r load_data}
library(data.table, quietly = TRUE)
rna <- fread("https://github.com/gabraham/flashpca/raw/master/data/DILGOM/rna.txt.gz")
metab <- fread("https://github.com/gabraham/flashpca/raw/master/data/DILGOM/metab.txt.gz")
rna <- as.matrix(rna)
metab <- as.matrix(metab)
dim(rna)
dim(metab)
```

We load the flashpcaR package as well as supporting packages for plotting. We
also use parallel computing with two cores to speed up the computation,
although this is not strictly necessary.
```{r libraries}
library(flashpcaR, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(ggVennDiagram, quietly = TRUE)
```

We can enable parallel computing to speed up the cross-validation.
```{r parallel,eval=FALSE}
library(doMC, quietly = TRUE)
registerDoMC(cores = 2)
```

Now we are ready to run SCCA in cross-validation on the data. This can take
some time depending on the size of the data, how many penalties we use in the
grid search, the number of dimensions, and the number of cross-validation folds.
We use parallelisation to make it a faster.

To save time, we will download a precomputed results from github:
```{r run_cv_scca,cache=TRUE}

# Cross-validation to select optimal penalties
set.seed(2329)
lambda1 <- seq(1e-2, 0.05, length = 30)
lambda2 <- seq(1e-5, 0.15, length = 20)

tmpdir <- tempdir()
f <- paste0(tmpdir, "/scca_cv.rda")
download.file("https://github.com/gabraham/flashpca/raw/master/data/DILGOM/scca_cv.rda", f)
load(f)
if (!exists("s1")) {
   s1 <- cv.scca(
      X = rna,
      Y = metab,
      ndim = 2,
      standx = "sd",
      standy = "sd",
      nfolds = 10,
      lambda1 = lambda1,
      lambda2 = lambda2,
      parallel = TRUE
   )
}
print(s1)
```

Now plot the cross-validated correlations for the 2D penalty grid

```{r plot_scca}
plot2d(s1, plot = TRUE)
```

Run SCCA on the entire dataset using the optimal penalties obtained earlier

```{r run_scca}

# Use the optimal penalties to run SCCA on all the data again
s2 <- scca(
   X = rna,
   Y = metab,
   stand = "sd",
   standy = "sd",
   ndim = 2,
   lambda1 = s1$best.lambda1,
   lambda2 = s1$best.lambda2,
   verbose = TRUE
)

# Extract the 1st canonical vectors
u1 <- s2$U[, 1]
names(u1) <- colnames(rna)
v1 <- s2$V[, 1]
names(v1) <- colnames(metab)
u1n0 <- u1[u1 != 0]
v1n0 <- v1[v1 != 0]

# Get the top variables with non-zero effect size
u1n0s <- sort(abs(u1n0), decreasing = TRUE)
v1n0s <- sort(abs(v1n0), decreasing = TRUE)
```

The selected gene expression probes (sorted by descending absolute weight)

```{r results_rna}
print(u1n0s)
```

The selected metabolites (sorted by descending absolute weight)

```{r results_metab}
print(v1n0s)
```

Now, we can map the Illumina gene expression probes to gene symbols, using Illumina's mapping information
for the Human-HT12 gene expression chip

```{r map_genes, cache=TRUE}
# Map the gene expression probe IDs to gene symbols
du1 <- data.table(
   Probe_Id = names(u1n0s),
   weight = u1n0s,
   rank = seq_along(u1n0s)
)
url <- "https://sapac.support.illumina.com/content/dam/illumina-support/documents/downloads/productfiles/humanht-12/v3/humanht-12_v3_0_r3_11283641_a_txt.zip"
tmpdir <- tempdir()
f <- paste0(tmpdir, basename(url))
if (!file.exists(f)) {
   download.file(url, f)
}
annot <- fread(unzip(f, "HumanHT-12_V3_0_R3_11283641_A.txt"), skip = 8, fill = TRUE)
du1[annot, on = "Probe_Id", Symbol := i.Symbol]
```

Finally, we can see whether the genes we have found map to the LL-module described earlier in [@inouye2010immune]:
```{r ll_module}
# The LL module in terms of Illumina probe IDs and gene symbols
LL.mod <- data.table(
   Symbol = c(
      "CPA3", "ENPP3", "FCER1A", "GATA2",
      "HDC", "HS.132563", "MS4A2", "MS4A3", "MS4A3", "SLC45A3", "SPRYD5"
   ),
   Probe_Id = c(
      "ILMN_1766551", "ILMN_1749131", "ILMN_1688423",
      "ILMN_2102670", "ILMN_1792323", "ILMN_1899034", "ILMN_1806721",
      "ILMN_1695530", "ILMN_1751625", "ILMN_1726114", "ILMN_1753648"
   )
)

LL.mod[, in.data := Probe_Id %in% colnames(rna)]

# Find whether the genes we find were in the LL module
# Note: possily not all LL modules genes will be represented
# in our result with non-zero weight
du1[, in.LL.module := FALSE]
du1[LL.mod, on = "Probe_Id", in.LL.module := TRUE]

print(du1)
```

We see that SCCA has identified a set of genes (probes) associated with metabolite levels, and that the top genes identified indeed correspond to the genes described in the LL module.

## References
